Competing Bandits:\\ The Perils of Exploration under Competition \thanks{This is a merged and final version of two conference papers, \citet{CompetingBandits-itcs18} and \citet{CompetingBandits-ec19}, with a unified and streamlined presentation.}

### 2020-07-19 journal paper/content/sec-bg.tex
typos / minor stylistic fixes
Minor edits.

\xhdr{Fundamentals.}
We are concerned with the following problem. There are $T$ rounds and $K$ \emph{arms} to choose from. In each round $t\in [T]$, the algorithm chooses an arm and receives a reward $r_t\in[0,1]$ for this arm, drawn from a fixed but unknown distribution. Algorithm's goal is to maximize the total reward.

@@

The three classes of algorithms perform very differently in terms of regret: adaptive-exploration algorithms are by far the best, greedy algorithms are by far the worst, and exploration-separating ones are in the middle. Adaptive-exploration algorithms achieve optimal regret rates:
    $\tildeO(\sqrt{KT})$
for all problem instances, and simultaneously a vastly improved regret rate of
    $O(\tfrac{K}{\Delta}\log T)$
for all problem instances with $\text{gap}\geq \Delta$, without knowing the $\Delta$ in advance
\citep{Lai-Robbins-85,bandits-ucb1,bandits-exp3}.%
\footnote{The \emph{gap} is the difference in mean reward between the best arm and the second-best arm. It is a standard notion of difficulty for MAB instances: instances with large gap are deemed  ``easy".}
Exploration-separating algorithms can only achieve regret $\tildeO(T^{2/3})$ across all problem instances. They can achieve the ``gap-dependent" regret rate stated above, but \emph{only} if they know the $\Delta$ in advance, and at the cost of a terrible regret $\Omega(\Delta T)$ for some other problem instances \citep{MechMAB-ec09}. Finally, the greedy algorithm is terrible on a wide variety of problem instances, in the sense that with constant probability it fails to try the best arm even once, and therefore suffers regret $\Omega(T)$
\citep[see Chapter 11.2 in][]{slivkins-MABbook}.

@@

Exploration-separating algorithms completely separate exploration and exploitation. Ahead of time, each round is either selected for exploration, in which case  the distribution over arms does not depend on the observed data, or it is assigned to exploitation, in which case the data from this round is discarded. The simplest approach, called \emph{Explore-First}, explores uniformly for a predetermined number of rounds, then chooses one arm for ``exploitation" and uses it from then on. A more refined approach, called \emph{Epsilon-Greedy}, explores uniformly in each round with a predetermined probability, and ``exploits" with the remaining probability. Both algorithms, and the associated $\tildeO(T^{2/3})$ regret bounds, have been ``folklore knowledge" for decades. The general definition of ``exploration-separating algorithms" and the associated lower bounds trace back to \citet{MechMAB-ec09}.%
\footnote{\citet{MechMAB-ec09} considers a closely related, but technically different setting, which can be easily ``translated" into ours (either as a corollary or as another application of the same proof technique). Detailing this is beyond the scope of this appendix.}

@@

@@

\xhdr{Instantaneous regret.}
Cumulative performance measures such as regret are not quite appropriate for our setting, as we need to characterize interactions in particular rounds. Instead, our theoretical results focus on \emph{Bayesian instantaneous regret} (\BIR), as defined in Section~\ref{sec:theory-prelims}. Recall that we posit a Bayesian prior on the \MRVs. In the notation of this appendix, the \BIR is simply:
\begin{align*}%\label{eq:bg-BIR-defn}
\BIR(t) := \E_{\text{prior}}\sbr{ \max_{a\in A} \mu_a - r_t}.
\end{align*}
\noindent Note that Bayesian regret (\ie regret in expectation over the prior) is precisely
\begin{align}\label{eq:bg-BReg}
  \BReg(T):=
    \E_{\text{prior}}\sbr{ T\cdot \max_{\text{arms $a$}}\mu_a - \sum_{t=1}^T r_t }
    = \sum_{t=1}^T \BIR(t).
\end{align}
We are primarily interested in how fast \BIR decreases with $t$, treating $K$ as a constant.

@@

\item Exploration-separating algorithms can achieve
    $\BIR(t) = \tildeO\rbr{t^{-1/3}}$
for all priors, \eg by using Epsilon-Greedy algorithm with exploration probability $\eps_t = t^{-1/3}$ in each round $t$.
In the typical scenario when
 $\BReg(t)\geq \Omega(t^{2/3})$,
the \BIR rate of $t^{-1/3}$  cannot be improved, in the following sense:
if
    $\BIR(t) = \tildeO\rbr{t^{-\gamma}}$
for all $t$, then $\gamma\geq \nicefrac{1}{3}$.

@@

This theoretical intuition is supported by our numerical simulations: see 
Figure~\ref{relative_rep_plots} and Appendix~\ref{app:isol}.

>>>>

\xhdr{Fundamentals.}
We are concerned with the following problem. There are $T$ rounds and $K$ \emph{arms} to choose from. In each round $t\in [T]$, the algorithm chooses an arm and receives a reward $r_t\in[0,1]$ for this arm, drawn from a fixed but unknown distribution.%
\footnote{\asedit{All ``negative" results (\ie lower bounds on regret) assume reward distributions with constant variance.}}
The algorithm's goal is to maximize the total reward.

@@

The three classes of algorithms perform very differently in terms of regret: adaptive-exploration algorithms are by far the best, greedy algorithms are by far the worst, and exploration-separating ones are in the middle. Adaptive-exploration algorithms achieve optimal regret rates:
    $\tildeO(\sqrt{KT})$
for all problem instances, and simultaneously a vastly improved regret rate of
    $O(\tfrac{K}{\Delta}\log T)$
for all problem instances with $\text{gap}\geq \Delta$ (``easy" instances), without knowing the $\Delta$ in advance
\citep{Lai-Robbins-85,bandits-ucb1,bandits-exp3}.%
\footnote{The \emph{gap} is the difference in mean reward between the best arm and the second-best arm.}
Exploration-separating algorithms can only achieve regret $\tildeO(T^{2/3})$ across all problem instances. They can achieve the ``gap-dependent" regret rate stated above, but \emph{only} if they know the $\Delta$ in advance, and with terrible regret $\Omega(\Delta T)$ for some other problem instances \citep{MechMAB-ec09}. Finally, the greedy algorithm is terrible on a wide variety of problem instances, in the sense that with constant probability it fails to try the best arm even once, and therefore suffers regret $\Omega(T)$
\citep[see Chapter 11.2 in][]{slivkins-MABbook}.

@@

Exploration-separating algorithms completely separate exploration and exploitation. Ahead of time, each round is either selected for exploration, in which case  the distribution over arms does not depend on the observed data, or it is assigned to exploitation, in which case the data from this round is discarded. The simplest approach, called \emph{Explore-First}, explores uniformly for a predetermined number of rounds, then chooses one arm for ``exploitation" and uses it from then on. A more refined approach, called \emph{Epsilon-Greedy}, explores uniformly in each round with a predetermined probability, and ``exploits" with the remaining probability. Both algorithms, and the associated $\tildeO(T^{2/3})$ regret bounds, have been ``folklore knowledge" for decades. The general definition of ``exploration-separating algorithms" and the associated lower bounds trace back to \citet{MechMAB-ec09}.%
\footnote{\citet{MechMAB-ec09} consider a closely related, but technically different setting, which can be easily ``translated" into ours (either as a corollary or as another application of the same proof technique). Detailing this is beyond the scope of this appendix.}

@@

\asedit{Some MAB algorithms are Bayesian, \eg Thompson Sampling,  in the sense that they input a prior on mean rewards, and attain strong Bayesian guarantees (in expectation over the prior) when the prior is correct. Such algorithms can also be initialized with some simple `fake' priors; in fact, Thompson Sampling satisfies the optimal regret bounds listed above.}

@@

\xhdr{Instantaneous regret.}
Cumulative performance measures such as regret are not quite appropriate for our setting, as we need to characterize interactions in particular rounds. Instead, our theoretical results focus on \emph{Bayesian instantaneous regret} (\BIR), as defined in Section~\ref{sec:theory-prelims}. Recall that we posit a Bayesian prior on the \MRVs. In the notation of this appendix, the \BIR is simply:
\begin{align*}%\label{eq:bg-BIR-defn}
\BIR(t) := \E_{\text{prior}}\sbr{ \max_{\text{arms $a$}} \mu_a - r_t}.
\end{align*}
\noindent Note that Bayesian regret (\ie regret in expectation over the prior) is precisely
\begin{align}\label{eq:bg-BReg}
  \BReg(T):=
    \E_{\text{prior}}\sbr{ T\cdot \max_{\text{arms $a$}}\mu_a - \sum_{t=1}^T r_t }
    = \sum_{t=1}^T \BIR(t).
\end{align}
We are primarily interested in how fast \BIR decreases with $t$, treating $K$ as a constant.

@@

\item Exploration-separating algorithms can achieve
    $\BIR(t) = \tildeO\rbr{t^{-1/3}}$
for all priors, \eg by using Epsilon-Greedy algorithm with exploration probability $\eps_t = t^{-1/3}$ in each round $t$.
In the typical scenario when
 $\BReg(t)\geq \Omega(t^{2/3})$,
the \BIR rate of $t^{-1/3}$  cannot be improved by \eqref{eq:bg-BReg}, in the following sense:
if
    $\BIR(t) = \tildeO\rbr{t^{-\gamma}}$
for all $t$, then $\gamma\geq \nicefrac{1}{3}$.

@@

\noindent This theoretical intuition is supported by our numerical simulations: see
Figure~\ref{relative_rep_plots} and Appendix~\ref{app:isol}.
